---
title: "Gini and the Iris Dataset"
author: "HGF"
date: "2024-11-18"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning =FALSE, message = FALSE)
```

# Inequality in plant data

Something introduced to me while studying for my PhD was the application of statistical models typically used for finance or demographics for biological data.

The thinking was that plant trials tend to take a large series of measurements, and that these plant populations and would demonstrate potentially large discrepancies under different treatments.

The best way to observe these differences is to see how homogeneous or indeed 'unequal' they are using these statistical models.

Here i'm going to apply the Gini Index to the Iris data set, and see if we can build any insights.


# Libraries & data

```{r}
library(gglorenz)
library(tidyverse)
library(ineq)
data(iris)
library(dplyr)
library(ggplot2)
library(rmarkdown)
```

## Gini

I'm interested in inequality. Typically one uses measurements of inequality for looking at 
distribution of wealth/GDP etc. Here I'll be applying it to the iris data set.

The higher the gini value (0-1), the more unequal the data is, and therefore, the less homogeneous.

```{r}
head(iris)
colnames(iris)

cols_numeric <-iris[sapply(iris, is.numeric)]

gini_vals <- sapply(cols_numeric, function(col) ineq(col, type = "Gini"))

print(gini_vals)
```

###  Not much to say!

That doesn't tell us much at all as we have a series of different species in the dataset.

Different species will have different morphological traits. Lets get a closer look at the homogeneity between species.

I'm going to actual make a function this time rather than have it as an anonymous one.

```{r}
cols_numeric <-iris[sapply(iris, is.numeric)]

gini_calc<-function(col){
    ineq(col, type = "Gini")
}

gini_species <- lapply(iris[1:4], function(col) {
    tapply(col, iris$Species, gini_calc)
    })

print(gini_species)

```

That's great, we can now see each species gini values for each morphological measurement, generally they all look very homogeneous. The only outlier looks like Setosa with its Petal.Width at 0.2, whereas all other values are <0.01.

Lets graph it next to see how it looks.

```{r}
ggplot(iris, aes(Petal.Width, color = Species), lwd = 1) +
stat_lorenz(desc = TRUE) +
geom_abline(linetype = "solid", color = "black", size=1) +
theme_classic()

```

Here we can see the *45^o* abline representing 0, or 'perfect equality'. The further away from this line our Petal Width for each species is, the less homogeneous (or 'higher inequality') that species petal width is. Essentially, from our Gini value of 0.20861789 for Setosa, and our graphing here with this Lorenz graph, we can say that the petal width of this species is much more unequal than other species. A measurement like this can be critical for manufacturing processes, agriculture/horticulture and a myriad of other industries as it identifies an element that falls outside the norm for producing consistent results.  