---
title: "OEE"
author: "HGF"
date: "2024-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries

```{r}
library(tidyverse)
```


#setwd("C:/Users/HGFea/Documents/GitHub/GitHub/HGfeatherz.github.io/projects")

## OEE

What is OEE? Overall equipment effectiveness (OEE) is a measure of how well a manufacturing operation is utilized (facilities, time and material) compared to its full potential, during the periods when it is scheduled to run. It identifies the percentage of manufacturing time that is truly productive. An OEE of 100% means that only good parts are produced (100% quality), at the maximum speed (100% performance), and without interruption (100% availability) [1] WIKI

## Using ML for OEE optimisation

Identifying anonmalies


## Example of Finding the OEE


```{r}
availability <- operating_time / planned_production_time
performance <- (ideal_cycle_time * total_pieces) / operating_time
quality <- good_pieces / total_pieces

oee <- availability * performance * quality
print(oee)
```

## Working with Open source data

Lets get going and bring in some data. Fortunately Kaggle offers a breadth of open source datasets. We'll use the Bosch Production Line Performance dataset.

#```{r}
files<-list.files(path = "projects/bosch-production-line-performance", pattern = "\\.csv$", recursive = TRUE)
base_path<-"projects/bosch-production-line-performance"

for (file in files) {
  full_path <-file.path(base_path, file)
  print(paste("processing:", full_path))
  data<-read.csv(full_path)
  print(head(data))
}

#```



