---
title: "Untitled"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning =FALSE, message = FALSE)
```

setwd("C:/Users/HGFea/Documents/Vs_code/phd")

# Introduction

Here is some of my data from the PhD I did a few years back, at the time I thought it would be great to write to whole thing in Markdown and Latex.

That was a mistake.

I'm now having a look at my old data to try and observe some  broad themes for the project.



# libraries 

```{r}
library(ggplot2)
library(readxl)
library(Gini)
library(dplyr)
library(tidyr)
library(tidyverse)
```

# import files

I have two file types to work with, one being catagorical for assessments, the other being a continous dataset.

```{r}
df<-read_excel("2021cont.xlsx")
df_cat<-read_excel("2021cat.xlsx")
```

# Exploring the dataset

It's been a few years since i've taken a look at this data, so lets have a quick peek and see whats going on.

### N counts for each observation

Due to the way I laid out this data, I strongly suspect there will be some missing values. I'll also see how many observations I had per 
treatment & experiment/trial. I'm doing this because the replicate number is missing from this data - i'll blame my past self for this later.
```{r}
summary(df)

na_counts <- sapply(df, function(x) sum(is.na(x)))
print(na_counts)

combo_counts <- df %>%
  count(exp, trtlong, meau, sub)

print(combo_counts)

```


As there are quite a few NA counts, mostly due to the data layout. Rather than mutating it right away, i'd like to see if I can filter it down to see some trends.

First thing i'll do is check out some of the unique values in certain cols. I split each observation type into a col called 'meau' (read: **measurement**), I also know that 'day' will be important
as certain observations occur at set time intervals (like harvest for instance).


```{r}
colnames(df)
unique_list <- lapply(df %>% select(day, meau, sub, amend), unique)
print(unique_list)
```

We can now see a list of days and meau's. Lets pick height and a certain day to build a quick graph.

We'll also filter for just 'pf' and 'peat' substrates, the 'pr' or 'Peat-reduced' wasn't included after the first few trials.

```{r}
unlist(unique(df$day))

# Lets filter out what we want to show!

df_height<-df%>%
        select(sub, meau, day, value, crop)%>%
        filter(sub == "peat" | sub == "pf", meau == 'height', day == "32")

# lets drop any NA values

df_height_clean <- df_height %>%
  filter(!is.na(value) & !is.infinite(value))

# I want to add a bit colour to my chart, so i'll need to factorise the 'sub' category first

df_height_clean2 <- df_height_clean %>%
  mutate(sub = factor(sub))


# and plotting

ggplot(df_height_clean2, aes(x = sub, y = value, fill = sub)) +
  geom_bar(position = "dodge", stat = "summary", fun = mean) +
  scale_fill_manual(values = c("peat" = "#6B873D", "pf" = "#068745")) +
  labs(x = "Treatment", y = "Mean Height") +
  ggtitle("Height at 32 Days")+
  facet_wrap(vars(crop))+
  theme_classic()

```

Looks like the peat substrate is greater in height at day 32 for both crop types. Lets add a bit more detail!

```{r}

df_summary <- df_height_clean2 %>%
  group_by(sub, crop) %>%
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    se_value = sd(value, na.rm = TRUE) / sqrt(n()),  
    n_count = n()  
  )


ggplot(df_summary, aes(x = sub, y = mean_value, fill = sub)) +
  geom_bar(position = "dodge", stat = "identity") +  
  scale_fill_manual(values = c("peat" = "#6B873D", "pf" = "#068745")) +
  geom_errorbar(aes(ymin = mean_value - se_value, ymax = mean_value + se_value), 
                position = position_dodge(width = 0.8), width = 0.25 ) + 
  geom_text(aes(label = round (mean_value, 2)), 
            position = position_dodge(width = 0.8), vjust = -4 ) +  
  labs(x = "Treatment", y = "Mean Height") +
  ggtitle("Height at 32 Days") +
  facet_wrap(vars(crop)) +
  theme_classic()


```

Here i've added error bars and a mean value for each bar. 

Now we've had a look at height at 32 days, lets have a quick look at the other observations in 'meau'

# Values within the treatments

```{r}

# filtering and summing

df_sum <- df %>%
    filter(sub == "peat" | sub == "pf")%>%
    group_by(meau, crop, sub) %>%
    summarise(mean_val = mean(value, na.rm = TRUE), .groups = "drop")

ggplot(df_sum, aes(x = mean_val, y = meau)) +
    geom_segment(aes(x = 0, y = meau, xend = mean_val, yend = meau)) + 
    geom_point(color = "#7e5fc4", size = 4) +
    theme_classic() +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(x = "Yield: POLY4 vs SP (%)", y = "P (ppm)") + 
    facet_wrap(vars(crop, sub))
```

 # lets get some interesting differences


```{r}

# calculating the means and dropping phygen (phygen is damage to leaves usualy from p&d, very little recorded so is being dropped)

df_sum <- df %>%
  filter(sub == "peat" | sub == "pf", meau != "phygen") %>%
  group_by(meau, crop, sub) %>%
  summarise(mean_val = mean(value, na.rm = TRUE), .groups = "drop")

# just a quick pivot

df_spread <- df_sum %>%
  pivot_wider(names_from = sub, values_from = mean_val) %>%
  mutate(
    pct_diff = (pf - peat) / peat * 100 
  )


#head(df_spread)

ggplot(df_spread, aes(x = pct_diff, y = meau)) +
  geom_segment(aes(x = 0, y = meau, xend = pct_diff, yend = meau), size = 1) + 
  geom_point(color = "#7e5fc4", size = 4) +
  theme_classic() +
  geom_vline(xintercept = 0, linetype = "dashed") +  
  labs(x = "Percentage Difference: PF vs PEAT (%)", y = "Measurement") + 
  ggtitle("Peat-Free vs Peat")
  facet_wrap(vars(crop))  


```

This graph is showing us the % difference for each measurement between peat and peat-free. As we can see, generally peat-free is performing poorly in most 
measurements, except for a rather intersting case in dry yield and rootlength.


## Old data
# Pivoting the dataset

Building two col values for the substrate types 'Peat' & 'pf'

```{r}
df_wide <- df %>%
    pivot_wider(
        names_from = sub,
        values_from = value,
        values_fn = list(value = mean)  # Replace `mean` with another function if needed
    )
```

```{r}
df_long <- df_cat %>%
    pivot_longer(
        cols = c(
            height,
            count, 
            freshyield,
            dryyield, 
            ld, 
            rootlength,
            potweight,
            moist,
            stemcount,
            phygen, 
            red, 
            green,
            blue
        ),
        names_to = "sub2",          # Corrected placement of `names_to`
        values_to = "value"         # Corrected placement of `values_to`
    )


```

# Generating a % difference between substrate types

```{r}
df_long <- df_cat %>%
    rename(meau_original = meau) %>%  
    pivot_longer(
        cols = c(height, count, freshyield, dryyield, ld, rootlength, potweight,
                 moist, stemcount, phygen, red, green, blue),
        names_to = "meau",           
        values_to = "value"         
    ) %>%
    pivot_wider(
        names_from = sub,            
        values_from = value,         
        values_fn = list(value = mean)  
    )

df_long2 <- df_long %>%
    mutate(diff_pf_peat = peat - pf)  


new_diff<- df_long2%>%
        group_by(meau,crop,)


```